Recovery of deleted deep sequencing data sheds more light on the early Wuhan SARS-CoV-2 epidemic

### 2021-06-27 paper/paper.tex
describe masking primer binding sites

\begin{table*}[b!]
{\small
\begin{tabular}{lrll}
\toprule
sample &  fraction sites called (21570-29550) &           patient group &                                        substitutions relative to proCoV2 \\
\midrule
A4 &                               0.9827 &        early outpatient &                                                                          none \\
C1 &                               0.9966 &        early outpatient &  G22081A (A=924, C=4, G=9), C28144T (C=6, T=1185), T29483G (C=1, G=45, T=1) \\
C2 &                               0.9962 &        early outpatient &                                                 C29095T (C=1, G=1, T=751)  \\
C9 &                               0.9536 &        early outpatient &                                  C28144T (C=3, T=823), G28514T (G=1, T=36)  \\
D9 &                               0.9585 &        early outpatient &                                                     C28144T (C=4, T=1653)  \\
D12 &                               0.9970 &        early outpatient &                                                     C28144T (C=8, T=2400)  \\
E1 &                               0.9759 &        early outpatient &                                                           C28144T (T=125)  \\
E5 &                               0.9758 &        early outpatient &              C24034T (A=5, C=3, T=74), T26729C (C=12), G28077C (C=142, G=4)  \\
E11 &                               0.9877 &        early outpatient &                                 C25460T (C=2, T=246), C28144T (C=1, T=412)  \\
F11 &                               0.9594 &        early outpatient &                             T25304A (A=9, T=1), C28144T (C=6, G=1, T=1328)  \\
G1 &                               0.9959 &        early outpatient &                                   none                                        \\
G11 &                               0.9677 &        early outpatient &                                  none                                         \\
H9 &                               0.9941 &        early outpatient &                                                     C28144T (C=2, T=1254)  \\
R11 &                               0.9987 &  hospital patient (Feb) &                               C21707T (T=401), C28144T (A=1, C=18, T=4265)  \\
\bottomrule
\end{tabular}
}
\caption{Samples for which the SARS-CoV-2 sequence could be called at $\ge$95\% of sites between 21,570 and 29,550, and the substitutions in this region relative to the putative SARS-CoV-2 progenitor proCoV2 inferred by \citet{kumar2021evolutionary}.
Numbers in parentheses after each substitution give the deep sequencing reads with each nucleotide identity.
\label{tab:mutations}
}
\end{table*}

@@

I aligned the recovered deep sequencing data to the SARS-CoV-2 genome using \texttt{minimap2}~\citep{li2018minimap2}, combining accessions for the same sample.
Figure~\ref{suppfig:coverage} shows the sequencing coverage for the 34 virus-positive early epidemic samples and the 16 hospitalized patient samples over the region of interest; a comparable plot for the whole genome is in Figure~\ref{suppfig:coverage_all}.

I called the consensus viral sequence for each sample at each site with coverage $\ge$3 and $>$80\% of the reads concurring on the nucleotide identity.
With these criteria, 13 of the early outpatient samples and 1 of the February hospitalized patient samples had sufficient coverage to call the consensus sequence at $>$95\% of the sites in the region of interest (Table~\ref{tab:mutations}), and for the remainder of this paper I focus on these high-coverage samples.
Table~\ref{tab:mutations} also shows the mutations in each sample relative to proCoV2, which is a putative progenitor of SARS-CoV-2 inferred by \citet{kumar2021evolutionary} that differs from the widely used Wuhan-Hu-1 reference sequence by three mutations (C8782T, C18060T, and T28144C).
Although requiring coverage of only $\ge$3 is relatively lenient, Table~\ref{tab:mutations} shows that all sites with mutations have coverage $\ge$10.
In addition, the mutations I called from the raw sequence data in Table~\ref{tab:mutations} concord with those mentioned in \citet{wang2020small}.

@@

\subsection{FASTQ files for SRR11313490 and SRR11313499}
I was not able to download SRA files for two runs, SRR11313490 and SRR11313499, from the Google Cloud.
After I posted the initial version of this pre-print, I was contacted by three different individuals (Andrew Fire -- Stanford University, Theo Sanderson -- Crick Institute, Francisco de Asis de Ribera -- DRASTIC Research) who had uncovered FASTQ or FASTA files for these two missing runs that appeared to have been downloaded from the SRA prior to the deletion of PRJNA612766.
For revised versions of the manuscript, I included analysis of the data for these two runs that I obtained from the links provided at \url{https://github.com/lifebit-ai/SARS-CoV-2/blob/master/assets/ucsc/aws_https_links.txt}.
Inclusion of data for these two runs did not appreciably change the results of the analysis relative to that in the original version of the pre-print, since both runs corresponded to low coverage samples for which meaningful viral genetic information could not be obtained.

@@

I processed the resulting alignments with \texttt{samtools} and \texttt{pysam} to determine the coverage at each site by aligned nucleotides with a quality score of at least 20.
These coverage plots are in Figure~\ref{suppfig:coverage} and Figure~\ref{suppfig:coverage_all}; the legends of these figures also link to interactive versions of the plots that enable zooming and mouseovers to get statistics for specific sites.
I called the consensus sequence at a site if this coverage was $\ge$3 and $>$80\% of the reads agreed on the identity.
These consensus sequences over the entire SARS-CoV-2 genome are available at \url{https://github.com/jbloom/SARS-CoV-2_PRJNA612766/raw/main/results/consensus/consensus_seqs.csv}; note that they are mostly \texttt{N} nucleotides since the sequencing approach of \citet{wang2020medRxiv} only covers part of the genome.

I only used the recovered consensus sequences in the downstream analyses if it was possible to call the consensus identity at $\ge$95\% of the sites in the region of interest from site 21,570 to 29,550.
These are the sequences listed in Table~\ref{tab:mutations}, and as described in that table, all mutation calls were at sites with coverage $\ge$10.
These sequences in the region of interest (21,570 to 29,550) are available at \url{https://github.com/jbloom/SARS-CoV-2_PRJNA612766/blob/main/results/recovered_seqs.fa}.

@@

\section{Acknowledgments}
I thank the citizens and scientists on Twitter who helped inspire this study and inform its background through their analyses and discussions of the early spread of SARS-CoV-2.
I thank the scientists and labs who contributed sequences used in this study to the GISAID database; the names of these scientists and labs are listed in the tables linked to in the Methods.
The scientific computing infrastructure used in this work was supported by the NIH Office of Research Infrastructure Programs under S10OD028685.
The author is an Investigator of the Howard Hughes Medical Institute.

>>>>

\begin{table*}[b!]
{\small
\begin{tabular}{lrll}
\toprule
sample &  fraction sites called (21570-29550) &           patient group &                                        substitutions relative to proCoV2 \\
\midrule
A4 &                               0.9827 &        early outpatient &                                                                          none \\
C1 &                               0.9966 &        early outpatient &  G22081A (A=924, C=4, G=9), C28144T (C=6, T=1185), T29483G (C=1, G=45, T=1) \\
C2 &                               0.9962 &        early outpatient &                                                 C29095T (C=1, G=1, T=751)  \\
C9 &                               0.9536 &        early outpatient &                                  C28144T (C=3, T=823), G28514T (G=1, T=36)  \\
D9 &                               0.9585 &        early outpatient &                                                     C28144T (C=4, T=1653)  \\
D12 &                               0.9970 &        early outpatient &                                                     C28144T (C=8, T=2400)  \\
E1 &                               0.9759 &        early outpatient &                                                           C28144T (T=125)  \\
E5 &                               0.9758 &        early outpatient &              C24034T (A=5, C=3, T=74), T26729C (C=12), G28077C (C=142, G=4)  \\
E11 &                               0.9877 &        early outpatient &                                 C25460T (C=2, T=246), C28144T (C=1, T=412)  \\
F11 &                               0.9594 &        early outpatient &                             T25304A (A=9, T=1), C28144T (C=6, G=1, T=1328)  \\
G1 &                               0.9959 &        early outpatient &                                   none                                        \\
G11 &                               0.9677 &        early outpatient &                                  none                                         \\
H9 &                               0.9941 &        early outpatient &                                                     C28144T (C=2, T=1254)  \\
R11 &                               0.9987 &  hospital patient (Feb) &                               C21707T (T=401), C28144T (A=1, C=18, T=4265)  \\
\bottomrule
\end{tabular}
}
\caption{Samples for which the SARS-CoV-2 sequence could be called at $\ge$90\% of sites between 21,570 and 29,550, and the substitutions in this region relative to the putative SARS-CoV-2 progenitor proCoV2 inferred by \citet{kumar2021evolutionary}.
Numbers in parentheses after each substitution give the deep sequencing reads with each nucleotide identity.
\label{tab:mutations}
}
\end{table*}

@@

I aligned the recovered deep sequencing data to the SARS-CoV-2 genome using \texttt{minimap2}~\citep{li2018minimap2}, combining accessions for the same sample, and masking regions that corresponded to the primer binding sites described in \citet{wang2020medRxiv}.
Figure~\ref{suppfig:coverage} shows the sequencing coverage for the 34 virus-positive early epidemic samples and the 16 hospitalized patient samples over the region of interest; a comparable plot for the whole genome is in Figure~\ref{suppfig:coverage_all}.

I called the consensus viral sequence for each sample at each site with coverage $\ge$3 and $>$80\% of the reads concurring on the nucleotide identity.
With these criteria, 13 of the early outpatient samples and 1 of the February hospitalized patient samples had sufficient coverage to call the consensus sequence at $>$90\% of the sites in the region of interest (Table~\ref{tab:mutations}), and for the remainder of this paper I focus on these high-coverage samples.
Table~\ref{tab:mutations} also shows the mutations in each sample relative to proCoV2, which is a putative progenitor of SARS-CoV-2 inferred by \citet{kumar2021evolutionary} that differs from the widely used Wuhan-Hu-1 reference sequence by three mutations (C8782T, C18060T, and T28144C).
Although requiring coverage of only $\ge$3 is relatively lenient, Table~\ref{tab:mutations} shows that all sites with mutations have coverage $\ge$10.
In addition, the mutations I called from the raw sequence data in Table~\ref{tab:mutations} concord with those mentioned in \citet{wang2020small}.

@@

\subsection{FASTQ files for SRR11313490 and SRR11313499}
I was not able to download SRA files for two runs, SRR11313490 and SRR11313499, from the Google Cloud.
After I posted the initial version of this pre-print, I was contacted by three different individuals (see Acknowledgments) who had uncovered FASTQ or FASTA files for these two missing runs that were downloaded from the SRA prior to the deletion of PRJNA612766.
For revised versions of the manuscript, I included analysis of the data for these two runs that I obtained from the links provided at \url{https://github.com/lifebit-ai/SARS-CoV-2/blob/master/assets/ucsc/aws_https_links.txt}.
Inclusion of data for these two runs did not appreciably change the results of the analysis relative to that in the original version of the pre-print, since both runs corresponded to low coverage samples for which meaningful viral genetic information could not be obtained.

@@

I processed the resulting alignments with \texttt{samtools} and \texttt{pysam} to determine the coverage at each site by aligned nucleotides with a quality score of at least 20.
I also masked (set to zero coverage) all sites overlapped by the primers used by \citet{wang2020medRxiv} to PCR the amplicons used for sequencing.
These coverage plots are in Figure~\ref{suppfig:coverage} and Figure~\ref{suppfig:coverage_all}; the legends of these figures also link to interactive versions of the plots that enable zooming and mouseovers to get statistics for specific sites.
I called the consensus sequence at a site if this coverage was $\ge$3 and $>$80\% of the reads agreed on the identity.
These consensus sequences over the entire SARS-CoV-2 genome are available at \url{https://github.com/jbloom/SARS-CoV-2_PRJNA612766/raw/main/results/consensus/consensus_seqs.csv}; note that they are mostly \texttt{N} nucleotides since the sequencing approach of \citet{wang2020medRxiv} only covers part of the genome.

I only used the recovered consensus sequences in the downstream analyses if it was possible to call the consensus identity at $\ge$90\% of the sites in the region of interest from site 21,570 to 29,550.
These are the sequences listed in Table~\ref{tab:mutations}, and as described in that table, all mutation calls were at sites with coverage $\ge$10.
These sequences in the region of interest (21,570 to 29,550) are available at \url{https://github.com/jbloom/SARS-CoV-2_PRJNA612766/blob/main/results/recovered_seqs.fa}.

@@

\section{Acknowledgments}
I thank the citizens and scientists on Twitter who helped inspire this study and inform its background through their analyses and discussions of the early spread of SARS-CoV-2.
I thank Brendan Larsen for pointing out that I failed to mask the primer binding site regions in the analysis described in the original pre-print (this has been fixed in this version).
I thank Andrew Fire (Stanford University), Theo Sanderson (Crick Institute), and Francisco de Asis de Ribera (DRASTIC Research) for finding locations of downloaded data for the two accessions (SRR11313490 and SRR11313499) that I could not download from the Google Cloud and so had to omit from the analysis described in the original pre-print (they are included in the analysis in this version).
I thank the scientists and labs who contributed sequences used in this study to the GISAID database; the names of these scientists and labs are listed in the tables linked to in the Methods.
The scientific computing infrastructure used in this work was supported by the NIH Office of Research Infrastructure Programs under S10OD028685.
The author is an Investigator of the Howard Hughes Medical Institute.
